---
title: "Regression and Other Stories: Vitals"
author: "Andrew Gelman, Jennifer Hill, Aki Vehtari"
date: "`r format(Sys.Date())`"
---
Predict weight. See Chapters 3, 9 and 10 in Regression and Other
Stories.

-------------


```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, error=FALSE, warning=FALSE, comment=NA)
```

**Load packages**

```{r }
library("rprojroot")
root<-has_dirname("ROS-Examples")$make_fix_file()
library("rstanarm")
```

**Load data**

```{r }
vitals_full <- read.csv(root("Vitals/data","vitals.csv"))
minidata <- vitals_full[, c("weight","height","female","ethnicity","exercise","smokenow")]
minidata$weight[minidata$weight>990]=NA;
ok <- apply(is.na(minidata), 1, sum) == 0
vitals <- minidata[ok,]
```

**Show some data**

```{r }
print(vitals[1:5,])
```

### Simulating uncertainty for linear predictors and predicted values
**Predict weight (in pounds) from height (in inches)**

The option `refresh = 0` supresses the default Stan sampling
progress output. This is useful for small data with fast
computation. For more complex models and bigger data, it can be
useful to see the progress.

```{r }
fit_1 <- stan_glm(weight ~ height, data=vitals, refresh = 0)
print(fit_1)
```

**Predict weight for 66 inches person

```{r }
coefs_1 <- coef(fit_1)
predicted_1 <- coefs_1[1] + coefs_1[2]*66
```

**Center heights**

```{r }
vitals$c_height <- vitals$height - 66
fit_2 <- stan_glm(weight ~ c_height, data=vitals, refresh = 0)
print(fit_2)
```

**Point prediction**

```{r }
new <- data.frame(c_height=4.0)
point_pred_2 <- predict(fit_2, newdata=new)
```

**Posterior simulations**

variation coming from posterior uncertainty in the coefficients

```{r }
linpred_2 <- posterior_linpred(fit_2, newdata=new)
hist(linpred_2)
```

**Posterior predictive simulations**

variation coming from posterior uncertainty in the coefficients and
predictive uncertainty

```{r }
postpred_2 <- posterior_predict(fit_2, newdata=new)
hist(postpred_2)
```

### Indicator variables
**Predict weight (in pounds) from height (in inches)**

```{r }
new <- data.frame(height=66)
pred <- posterior_predict(fit_1, newdata=new)
cat("Predicted weight for a 66-inch-tall person is", round(mean(pred)), "pounds with a sd of", round(sd(pred)), "\n")
```

**Including a binary variable in a regression**

```{r }
fit_3 <- stan_glm(weight ~ c_height + female, data=vitals, refresh = 0)
print(fit_3)
new <- data.frame(c_height=4, female=1)
pred <- posterior_predict(fit_3, newdata=new)
cat("Predicted weight for a 70-inch-tall female is", round(mean(pred)), "pounds with a sd of", round(sd(pred)), "\n")
```

**Using indicator variables for multiple levels of a categorical predictor**<br/>
Include ethnicity in the regression as a factor

```{r }
fit_4 <- stan_glm(weight ~ c_height + female + factor(ethnicity),
                  data=vitals, refresh = 0)
print(fit_4)
```

Choose the baseline category by setting the levels

```{r }
vitals$eth <- factor(vitals$ethnicity,
  levels=c("white", "black", "hispanic", "other"))
fit_5 <- stan_glm(weight ~ c_height + female + eth, data=vitals, refresh = 0)
print(fit_5)
```

Alternatively create indicators for the four ethnic groups directly:

```{r }
vitals$eth_white <- ifelse(vitals$ethnicity=="white", 1, 0)
vitals$eth_black <- ifelse(vitals$ethnicity=="black", 1, 0)
vitals$eth_hispanic <- ifelse(vitals$ethnicity=="hispanic", 1, 0)
vitals$eth_other <- ifelse(vitals$ethnicity=="other", 1, 0)
fit_6 <- stan_glm(weight ~ c_height + female + eth_black + eth_hispanic + eth_other,
                  data=vitals, refresh = 0)
print(fit_6)
```

