---
title: "Regression and Other Stories: Earnings"
author: "Andrew Gelman, Jennifer Hill, Aki Vehtari"
date: "`r format(Sys.Date())`"
---
Predict respondents' yearly earnings using survey data from
1990. See Chapters 6, 9 and 12 in Regression and Other Stories.

-------------


```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, error=FALSE, warning=FALSE, comment=NA)
# switch this to TRUE to save figures in separate files
savefigs <- FALSE
```

**Load packages**

```{r }
library("rprojroot")
root<-has_dirname("ROS-Examples")$make_fix_file()
library("rstanarm")
options(mc.cores = parallel::detectCores())
library("ggplot2")
library("bayesplot")
theme_set(bayesplot::theme_default(base_family = "sans"))
```
```{r eval=FALSE, include=FALSE}
# grayscale figures for the book
if (savefigs) color_scheme_set(scheme = "gray")
```

Set random seed for reproducability

```{r }
SEED <- 7783
```

**Load data**

```{r }
earnings_all <- read.csv(root("Earnings/data","earnings.csv"))
earnings_all$positive <- earnings_all$earn > 0
# scale earnings to thousands of dollars
earnings_all$earnk <- earnings_all$earn/1000 
# only non-zero earnings
earnings <- earnings_all[earnings_all$positive, ]
n <- nrow(earnings)
height_jitter_add <- runif(n, -.2, .2)
```

### Normal linear regression
**Model on 1k dollars scale**

The option `refresh = 0` supresses the default Stan sampling
progress output. This is useful for small data with fast
computation. For more complex models and bigger data, it can be
useful to see the progress.

```{r }
fit_1 <- stan_glm(earnk ~ height, data = earnings,
                  seed = SEED, refresh = 0)
print(fit_1)
```

for plotting scale back to dollar scale

```{r }
coef1 <- coef(fit_1)*1000
```

**Plot linear model**

```{r eval=FALSE, include=FALSE}
if (savefigs) pdf(root("Earnings/figs","heights1a.pdf"), height=8.5, width=11)
```
```{r }
par(mar=c(6,6,4,2)+.1)
plot(earnings$height + height_jitter_add, earnings$earn, xlab="height", ylab="earnings",
     cex=.8, cex.lab=3, pch=20, cex.axis=3, yaxt="n", mgp=c(4,1.5,0),
     col="gray10",  cex.main=3, main="Fitted linear model")
abline(coef1, lwd=2)
abline(0,0,col="gray")
axis(2, c(0,100000,200000), c("0","100000","200000"), mgp=c(4,1.1,0),cex.axis=3)
```
```{r eval=FALSE, include=FALSE}
if (savefigs) dev.off()
```

**Plot linear model, ggplot version**

```{r }
gg_earnings <- ggplot(earnings, aes(x = height + height_jitter_add, y = earn)) +
  geom_point(alpha = 0.75) +
  geom_hline(yintercept = 0, color = "darkgray") +
  geom_abline(intercept = coef1[1], slope = coef1[2], size = 1) +
  labs(x = "height", y = "earnings",
       title = "Fitted linear model")
gg_earnings
```

**Plot extrapolation**

```{r eval=FALSE, include=FALSE}
if (savefigs) pdf(root("Earnings/figs","heights1b.pdf"), height=8.5, width=11)
```
```{r }
par(mar=c(6,6,4,2)+.1)
plot(xlim=c(0,max(earnings$height)), ylim=c(-70000,200000), earnings$height + height_jitter_add, earnings$earn, xlab="height", ylab="earnings", cex=.8, cex.lab=3, pch=20, cex.axis=3, yaxt="n", mgp=c(4,1.5,0), col="gray10", cex.main=3, main="Extrapolation")
abline(coef1, lwd=2)
abline(0,0,col="gray")
intercept <- coef1[1]
axis(2, c(intercept,0,100000,200000), c(round(intercept,-2),"0","100000","200000"), mgp=c(4,1.1,0),cex.axis=3)
```
```{r eval=FALSE, include=FALSE}
if (savefigs) dev.off()
```

**Plot extrapolation, ggplot version, modifying the gg_earnings object
we already created**

```{r }
gg_earnings +
  ylim(-70000, 200000) +
  xlim(0, 80) +
  labs(title = "Extrapolation")
```

**Include male/female**

```{r results='hide'}
fit_2 <- stan_glm(earnk ~ height + male, data = earnings,
                  seed = SEED)
```
```{r }
print(fit_2)
```

for plotting scale back to dollar scale

```{r }
coef2 <- coef(fit_2)*1000
```

**Plot linear model with male/female**

```{r eval=FALSE, include=FALSE}
if (savefigs) pdf(root("Earnings/figs","heights2.pdf"), height=8.5, width=11)
```
```{r }
par(mar=c(6,6,5,2)+.1)
plot(range(earnings$height), range(predict(fit_2)*1000), type="n", xlab="height", ylab="predicted earnings", cex=.8, cex.lab=3, pch=20, cex.axis=3, mgp=c(4,1.5,0), yaxt="n", col="gray10",
      cex.main=3, main="Fitted regression, displayed as\nseparate lines for men and women", bty="l")
axis(2, c(20000,30000), cex.axis=3)
abline(coef2[1], coef2[2], col="red", lwd=2)
text(68, coef2[1] + coef2[2]*65, "women:\ny = -11 000 + 450x", cex=3, adj=0, col="red")
abline(coef2[1]+coef2[3], coef2[2], col="blue", lwd=2)
text(68, coef2[1]+coef2[3] + coef2[2]*65, "men:\ny = -2 000 + 450x", cex=3, adj=0, col="blue")
```
```{r eval=FALSE, include=FALSE}
if (savefigs) dev.off()
```

**Include male/female, ggplot version**

```{r }
ggplot(earnings, aes(height, earn)) +
  geom_blank() +
  geom_abline(
    intercept = c(coef2[1], coef2[1] + coef2[3]),
    slope = coef2[2],
    color = c("red", "blue")
  ) +
  coord_cartesian(
    ylim = range(predict(fit_2)*1000),
    xlim = range(earnings$height)
  ) +
  annotate(
    "text",
    x = c(68, 68),
    y = c(coef2[1] + coef2[2] * 65, coef2[1] + coef2[3] + coef2[2] * 65),
    label = c("women:\ny = -11 000 + 450x", "men:\ny = -2 000 + 450x"),
    color = c("red", "blue"),
    size = 5, hjust = 0
  ) +
  labs(
    x = "height",
    y = "predicted earnings",
    title = "Fitted regression, displayed as\nseparate lines for men and women"
  )
```

**Include interaction**

```{r results='hide'}
fit_3 <- stan_glm(earnk ~ height + male + height:male, data = earnings,
                  seed = SEED, refresh = 0)
```
```{r }
print(fit_3)
```

for plotting scale back to dollar scale

```{r }
coef3 <- coef(fit_3)*1000
```

**Plot linear model with interaction**

```{r eval=FALSE, include=FALSE}
if (savefigs) pdf(root("Earnings/figs","heights3.pdf"), height=8.5, width=11)
```
```{r }
par(mar=c(6,6,5,2)+.1)
plot(range(earnings$height), range(predict(fit_3)*1000), type="n", xlab="height", ylab="predicted earnings", cex=.8, cex.lab=3, pch=20, cex.axis=3, mgp=c(4,1.5,0), yaxt="n", col="gray10", cex.main=3, main="Fitted regression with interactions,\nseparate lines for men and women", bty="l")
axis(2, c(20000,30000), cex.axis=3)
abline(coef3[1], coef3[2], col="red", lwd=2)
text(62, coef3[1] + coef3[2]*80, "women:\ny = -7 000 + 180x", cex=3, adj=0, col="red")
abline(coef3[1]+coef3[3], coef3[2]+coef3[4], col="blue", lwd=2)
text(68, coef3[1]+coef3[3] + (coef3[2]+coef3[4])*66, "men:\ny = -22 000 + 740x", cex=3, adj=0, col="blue")
```
```{r eval=FALSE, include=FALSE}
if (savefigs) dev.off()
```

**Include interaction, ggplot version**

```{r }
ggplot(earnings, aes(height, earn)) +
  geom_blank() +
  geom_abline(
    intercept = c(coef3[1], coef3[1] + coef3[3]),
    slope = c(coef3[2], coef3[2] + coef3[4]),
    color = c("red", "blue")
  ) +
  coord_cartesian(
    ylim = range(predict(fit_3)*1000),
    xlim = range(earnings$height)
  ) +
  annotate(
    "text",
    x = c(62, 68),
    y = c(coef3[1] + coef3[2] * 80, coef3[1]+coef3[3] + (coef3[2]+coef3[4])*66),
    label = c("women:\ny = -7 000 + 180x", "men:\ny = -22 000 + 740x"),
    color = c("red", "blue"),
    size = 5, hjust = 0
  ) +
  labs(
    x = "height",
    y = "predicted earnings",
    title = "Fitted regression with interactions,\nseparate lines for men and women"
  )
```

### Linear regression on log scale
**Models on log scale**

```{r }
earnings$log_earn <- log(earnings$earn)
```
```{r results='hide'}
logmodel_1 <- stan_glm(log_earn ~ height, data = earnings,
                       seed = SEED, refresh = 0)
```
```{r }
print(logmodel_1, digits=2)
```

**Model on log10 scale**

```{r }
earnings$log10_earn <- log10(earnings$earn)
```
```{r results='hide'}
log10model_1 <- stan_glm(log10_earn ~ height, data = earnings,
                         seed = SEED, refresh = 0)
```
```{r }
print(log10model_1, digits=3)
```

**Model on log scale with two predictors**

```{r results='hide'}
logmodel_2 <- stan_glm(log_earn ~ height + male, data = earnings,
                       seed = SEED, refresh = 0)
```
```{r }
print(logmodel_2, digits=2)
```

**Model on log scale for the target and one predictor**

```{r }
earnings$log_height <- log(earnings$height)
```
```{r results='hide'}
loglogmodel_2 <- stan_glm(log_earn ~ log_height + male, data = earnings,
                          seed = SEED, refresh = 0)
```
```{r }
print(loglogmodel_2, digits=2)
```

**Model on log scale with two predictors and interaction**

```{r }
logmodel_3 <- stan_glm(log_earn ~ height + male + height:male, data = earnings,
                       seed = SEED, refresh = 0)
```
```{r }
print(logmodel_3, digits=2)
```

**Model on log scale with standardized interaction**

```{r }
earnings$z_height <- with(earnings, (height - mean(height))/sd(height))
```
```{r results='hide'}
logmodel_3a <- stan_glm(log_earn ~ z_height + male + z_height:male,
                        data = earnings, seed = SEED, refresh = 0)
```
```{r }
print(logmodel_3a, digits=2)
```

### Uncertainty
**Posterior uncertainty for log model**

```{r }
sims <- as.matrix(logmodel_2)
n_sims <- nrow(sims)
```

**Plot posterior draws of linear model on log scale**

```{r eval=FALSE, include=FALSE}
postscript(root("Earnings/figs","heights.log1a.ps"), horizontal=TRUE)
```
```{r }
par(mar=c(6,6,4,2)+.1)
plot(earnings$height + runif(n,-.2,.2), earnings$log_earn, xlab="height", ylab="log (earnings)", cex=.8, cex.lab=3, pch=20, cex.axis=3, yaxt="n", mgp=c(4,1.5,0), col="gray10",
      cex.main=3, main="Log regression, plotted on log scale")
axis(2, seq(6,12,2), mgp=c(4,1.1,0),cex.axis=3)
sims_display <- sample(n_sims, 10)
for (i in sims_display){
  curve(sims[i,1] + sims[i,2]*x, lwd=0.5, col="gray30", add=TRUE)
}
curve(coef(logmodel_2)[1] + coef(logmodel_2)[2]*x, add=TRUE)
```
```{r eval=FALSE, include=FALSE}
if (savefigs) dev.off()
```

**Plot posterior draws of linear model on log scale, ggplot version**

```{r }
sims_display <- sample(n_sims, 10)
ggplot(earnings, aes(height, log_earn)) +
  geom_jitter(height = 0, width = 0.25) +
  geom_abline(
    intercept = sims[sims_display, 1],
    slope = sims[sims_display, 2],
    color = "darkgray"
  ) +
  geom_abline(
    intercept = coef(logmodel_2)[1],
    slope = coef(logmodel_2)[2]
  ) +
  labs(
    x = "height",
    y = "log(earnings)",
    title = "Log regression, plotted on log scale"
  )
```

### Posterior predictive checking
**Posterior predictive checking for model in linear scale**

```{r }
yrep_1 <- posterior_predict(fit_1)
n_sims <- nrow(yrep_1)
sims_display <- sample(n_sims, 100)
ppc_1 <- ppc_dens_overlay(earnings$earnk, yrep_1[sims_display,])
```

**Posterior predictive checking for model in log scale**

```{r }
yrep_log_1 <- posterior_predict(logmodel_1)
n_sims <- nrow(yrep_log_1)
sims_display <- sample(n_sims, 100)
ppc_log_1 <- ppc_dens_overlay(earnings$log_earn, yrep_log_1[sims_display,])
(bpg <- bayesplot_grid(
  ppc_1, ppc_log_1,
  grid_args = list(ncol = 2),
  titles = c("earn", "log(earn)")
))
```
```{r eval=FALSE, include=FALSE}
ggsave(root("Earnings/figs","earnings_ppc.pdf"), bpg, height=3, width=9)
```

**Posterior predictive checking for model in linear scale**

```{r }
yrep_2 <- posterior_predict(fit_2)
n_sims <- nrow(yrep_2)
sims_display <- sample(n_sims, 100)
ppc_dens_overlay(earnings$earnk, yrep_2[sims_display,])
```

**Posterior predictive checking for model in log scale**

```{r }
yrep_log_2 <- posterior_predict(logmodel_2)
n_sims <- nrow(yrep_log_2)
sims_display <- sample(n_sims, 100)
ppc_dens_overlay(earnings$log_earn, yrep_log_2[sims_display,])
```

**Posterior predictive checking for model in log-log scale**

```{r }
yrep_loglog_2 <- posterior_predict(loglogmodel_2)
n_sims <- nrow(yrep_loglog_2)
sims_display <- sample(n_sims, 100)
ppc_dens_overlay(earnings$log_earn, yrep_loglog_2[sims_display,])
```

