---
title: "Regression and Other Stories: RiskyBehavior"
author: "Andrew Gelman, Jennifer Hill, Aki Vehtari"
date: "`r format(Sys.Date())`"
---

Data from a randomized trial targeting couples at high risk of HIV
infection. The intervention provided counseling sessions regarding
practices that could reduce their likelihood of contracting
HIV. Couples were randomized either to a control group, a group in
which just the woman participated, or a group in which both members
of the couple participated. One of the outcomes examined after
three months was “number of unprotected sex acts.”. See Chapter 15
in Regression and Other Stories.

-------------


```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, error=FALSE, warning=FALSE, comment=NA)
# switch this to TRUE to save figures in separate files
savefigs <- FALSE
```

**Load packages**

```{r }
library("rprojroot")
root<-has_dirname("ROS-Examples")$make_fix_file()
```

**Load data**

```{r }
risky <- read.csv(root("RiskyBehavior/data","risky.csv"))
head(risky)
```

