---
title: "Regression and Other Stories: Newcomb"
author: "Andrew Gelman, Jennifer Hill, Aki Vehtari"
date: "`r format(Sys.Date())`"
---
Posterior predictive checking of Normal model for Newcomb's speed of light data

-------------


```{r include=FALSE}
# switch this to TRUE to save figures in separate files
savefigs <- FALSE
```

**Load packages**

```{r setup, message=FALSE, error=FALSE, warning=FALSE}
library("rprojroot")
root<-has_dirname("RAOS-Examples")$make_fix_file()
library("rstanarm")
options(mc.cores = parallel::detectCores())
library("ggplot2")
library("bayesplot")
theme_set(bayesplot::theme_default(base_family = "sans"))
color_scheme_set(scheme = "gray")
```

**Data**<br>
Simon Newcomb's measurements of the speed of light, from Stigler
(1977).  The data are recorded as deviations from $24,\!800$
nanoseconds.

```{r }
newcomb <- data.frame(y=c(28,26,33,24,34,-44,27,16,40,-2,29,22,24,21,25,30,
                          23,29,31,19,24,20,36,32,36,28,25,21,28,29,37,25,28,
                          26,30,32,36,26,30,22, 36,23,27,27,28,27,31,27,26,
                          33,26,32,32,24,39,28,24,25,32,25,29,27,28,29,16,23))
```

**Fit a regression model with just the intercept term**

```{r }
fit <- stan_glm(y ~ 1, data=newcomb)
```

**Simulate from the predictive distribution**

```{r }
sims <- as.matrix(fit)
n_sims <- nrow(sims)
n <- length(newcomb$y)
y_rep <- array(NA, c(n_sims, n))
for (s in 1:n_sims){
    y_rep[s,] <- rnorm(n, sims[s,1], sims[s,2])
}
```

**Plot histogram of 20 replicates**

```{r }
par(mfrow=c(5,4), mar=rep(2,4))
for (s in sample(n_sims, 20)) {
    hist(y_rep[s,], main=NULL, ylab="", xlab="", yaxt="n")
}
```

**Simulate using built-in function**

```{r }
y_rep <- posterior_predict(fit)
```

**Plot data and 19 replications using built-in function**

```{r eval=FALSE, include=FALSE}
if (savefigs) pdf(root("Newcomb/figs","newcomb_ppc_hist.pdf"), height=3, width=9)
```
```{r }
ppc_hist(newcomb$y, y_rep[1:19, ])
```
```{r eval=FALSE, include=FALSE}
if (savefigs) dev.off()
```

**Plot test statistic for data and replicates**

```{r }
Test <- function (y){
    min(y)
}
test_rep <- apply(y_rep, 1, Test)
hist(test_rep, xlim=range(Test(newcomb$y), test_rep))
lines(rep(Test(newcomb$y),2), c(0,n_sims))
```

**Plot test statistic for data and replicates using built-in function**

```{r eval=FALSE, include=FALSE}
if (savefigs) pdf(root("Newcomb/figs","newcomb_ppc_stat.pdf"), height=3, width=6)
```
```{r }
ppc_stat(newcomb$y, y_rep, stat = "min")
```
```{r eval=FALSE, include=FALSE}
if (savefigs) dev.off()
```

